---
title: "bees"
editor_options: 
  chunk_output_type: console
---

data download here: https://www.kaggle.com/datasets/m000sey/save-the-honey-bees/
some helpful eco info: https://www.ars.usda.gov/oc/br/ccd/index/
all reports here: https://usda.library.cornell.edu/concern/publications/rn301137d?locale=en
definitions here: https://downloads.usda.library.cornell.edu/usda-esmis/files/rn301137d/4m90gc28p/gq67m7401/hcny0823.pdf

- `num_colonies`: colonies counted at the start of the quarter
- `max_colonies`: `num_colonies` + all colonies that moved into the state that quarter (can't be used to calculate a national level of max colonies since colonies are counted in every state they were in during the quarter i.e. there is duplication)
- `percent_lost`: number of colonies lost / `max_colonies`

```{r}
library(tidyverse)

bees <- read_csv(here::here("bees", "save_the_bees.csv"))

bees_wrangled <- bees |> 
  mutate(quarter_months = case_when(
    quarter == 1 ~ "Jan - Mar",
    quarter == 2 ~ "Apr - Jun",
    quarter == 3 ~ "Jul - Sep",
    quarter == 4 ~ "Oct - Dec"
  )) |> 
  mutate(quarter_months = as.factor(quarter_months)) |> 
  
  
```


```{r}
bees |> 
  group_by(year, quarter) |> 
  ggplot(aes(x = state, y = varroa_mites)) +
  geom_col()
```

