---
title: "fracking"
editor_options: 
  chunk_output_type: console
---

# Fracking

- <https://www.usgs.gov/news/national-news-release/water-used-hydraulic-fracturing-varies-widely-across-united-states>
- <https://www.abc27.com/pennsylvania/12-states-where-fracking-is-most-prevalent-where-does-pa-stand/#:~:text=1.,Texas&text=Texas%20is%20situated%20over%20parts,gas%20anywhere%20in%20the%20U.S.>
- <https://www.data-is-plural.com/archive/2023-09-27-edition/>
- <https://fracfocus.org/data-download>

`TotalBaseWaterVolume`: The total volume of water used as a carrier fluid for the hydraulic fracturing job (in gallons).

### setup
```{r}
# load libraries
library(tidyverse)
library(janitor)
library(ggridges)

# read in data
fracking <- read_csv(here::here("fracking", "data", "registryupload_1.csv"))
```

## cleaning
```{r}
# clean data -----------------
fracking_clean <- fracking |> 
  
  # clean column names ----
  janitor::clean_names() |> 
  
  # clean up dates ----
  mutate(job_start_date = str_remove(job_start_date, " AM")) |> 
  mutate(datetime_start = mdy_hms(job_start_date)) |> 
  mutate(year = year(datetime_start)) |> 

  # select relevant cols ----
  select(datetime_start, year, state_name, county_name, well_name, total_base_water_volume) |> 
  
  # filter out non-state names ----
  filter(!state_name %in% c("Beaver", "Beckham", "Harper", "Hemphill", "Midland", "Red River", "Roosevelt", "Rusk", "State", "WARD")) |> 
  
  # rename state_name to something shorter for typing with case_when ----
  rename(sn = state_name) |> 
  
  # make all words title case ----
  mutate(sn = str_to_title(sn)) |> 
  
  # fix misspelled state names ----
  mutate(sn = case_when(
    sn == "Colordao" ~ "Colorado",
    sn == "Loiusiana" ~ "Louisiana",
    sn == "Louisianna" ~ "Louisiana",
    sn == "Lousiana" ~ "Louisiana",
    sn == "New Mexcio" ~ "New Mexico",
    sn == "Norh Dakota" ~ "North Dakota",
    sn == "Norht Dakota" ~ "North Dakota",
    sn == "North  Dakota" ~ "North Dakota",
    sn == "North Dakata" ~ "North Dakota",
    sn == "North Dakotta" ~ "North Dakota",
    sn == "Noth Dakota" ~ "North Dakota",
    sn == "Pennslvania" ~ "Pennsylvania",
    sn == "Pennsylavania" ~ "Pennsylvania",
    sn == "Pennsylvanya" ~ "Pennsylvania",
    sn == "Penssylvania" ~ "Pennsylvania",
    sn == "Texasa" ~ "Texas",
    sn == "Texs" ~ "Texas", 
    sn == "West Viginia" ~ "West Virginia",
    sn == "Wyominng" ~ "Wyoming", 
    TRUE ~ sn
  )) |> 
  
  # remove rows that have a ? mark ----
  filter(!str_detect(sn, "\\?")) |> 
  
  # make all works uppercase (so that we can covert abbreviations to state names) ----
  mutate(sn = str_to_upper(sn)) |> 
  
  # mutate abbrs to full state names ----
  mutate(sn = ifelse(test = str_length(sn) == 2, 
                     yes = usdata::abbr2state(sn),
                     no = sn)) |> 
  
  # make all words title case again ----
  mutate(sn = str_to_title(sn)) |> 
  
  # create a column of just state abbrs ----
  mutate(state_abb = usdata::state2abbr(sn)) |> 
  
  # rename sn to state_name again for clarity ----
  rename(state_name = sn, total_base_water_volume_gal = total_base_water_volume) |> 
  
  # move state_abb col after state_name col
  relocate(state_abb, .after = state_name) |> 
  
  # convert state_name & state_abb from string to factor ----
  mutate(state_name = as.factor(state_name),
         state_abb = as.factor(state_abb)) |> 
  
  # remove obs that don't have a measurement for total_base_water_volume (NA)
  drop_na(total_base_water_volume_gal)
```


-----

```{r}
fracking_clean |> 
  filter(state_name %in% c("Texas", "Colorado", "North Dakota")) |> 
  filter(year == 2015) |> 
  filter(total_base_water_volume_gal < 70000000) |>
  group_by(state_name) |> 
  mutate(state_name = fct_relevel(state_name, c("Texas", "Colorado", "North Dakota"))) |> 
  ggplot(aes(x = state_name, y = total_base_water_volume_gal, fill = state_name)) +
  geom_violin(alpha = 0.5) +
  scale_y_continuous(labels = scales::label_comma()) +
  viridis::scale_fill_viridis(discrete = TRUE) +
  labs(x = "State", y = "Total water volume used per fracking job (gallons)") +
  theme(
    legend.position = "none"
  )
```


-----



## State land area dataframe
```{r}
state_areas <- data.frame(state_name = usa::state.name, 
                          land_area_sq_mi = usa::state.area)
```

## Join dfs by state_name to add areas
```{r}
fracking_areas <- left_join(fracking_clean, state_areas)
```

## Subet (top 3 fracking states in year 2015 (last year with available data))
```{r}
fracking_top3_2015 <- fracking_areas |> 
  filter(state_name %in% c("Texas", "Colorado", "North Dakota")) |> 
  filter(year == 2015) |> 
  group_by(state_name) |> 
  mutate(state_name = fct_relevel(state_name, c("Texas", "Colorado", "North Dakota")))
```

## HISTOGRAM (TX, CO, ND in 2015)
```{r}
ggplot(fracking_top3_2015, aes(x = total_base_water_volume_gal, fill = state_name)) +
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 2500000) +
  viridis::scale_fill_viridis(discrete = TRUE) +
  labs(x = "Total water volume used per fracking job (gallons)",
       y = "Frequency",
       fill = "State") +
  scale_x_continuous(labels = scales::label_comma())
```

## DENSITY PLOT (TX, CO, ND in 2015)
```{r}
ggplot(fracking_top3_2015, aes(x = total_base_water_volume_gal, fill = state_name)) +
  geom_density(position = "identity", alpha = 0.5) +
  viridis::scale_fill_viridis(discrete = TRUE) +
  labs(x = "Total water volume used per fracking job (gallons)",
       y = "Frequency",
       fill = "State") +
  scale_x_continuous(labels = scales::label_comma())
```

## BOX PLOT (TX, CO, ND in 2015)
```{r}
ggplot(fracking_top3_2015, aes(x = state_name, y = total_base_water_volume_gal, fill = state_name)) +
  geom_boxplot(alpha = 0.5) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = "State", y = "Total water volume used per fracking job (gallons)") +
  theme(
    legend.position = "none"
  )
```

## VIOLIN PLOT (TX, CO, ND in 2015)
```{r}
ggplot(fracking_top3_2015, aes(x = state_name, y = total_base_water_volume_gal, fill = state_name)) +
  geom_violin(alpha = 0.5) +
  scale_y_continuous(labels = scales::label_comma()) +
  viridis::scale_fill_viridis(discrete = TRUE) +
  labs(x = "State", y = "Total water volume used per fracking job (gallons)") +
  theme(
    legend.position = "none"
  )
```

## RIDGELINE PLOT (TX, CO, ND in 2015)
```{r}
ggplot(fracking_top3_2015, aes(x = total_base_water_volume_gal, y = state_name, fill = state_name)) +
  ggridges::geom_density_ridges(alpha = 0.5) +
  scale_y_continuous(labels = scales::label_comma()) +
  viridis::scale_fill_viridis(discrete = FALSE) +
  labs(x = "State", y = "Total water volume used per fracking job (gallons)") +
  theme(legend.position = "none")
```

## SCATTER
```{r}
ggplot(fracking_top3_2015, aes(x = state_name, y = total_base_water_volume_gal, fill = state_name)) +
  geom_point(alpha = 0.5, position = "jitter") +
  scale_y_continuous(labels = scales::label_comma()) +
  viridis::scale_fill_viridis(discrete = TRUE) +
  labs(x = "State", y = "Total water volume used per fracking job (gallons)") +
  theme(
    legend.position = "none"
  )
```





